Evaluating the Capabilities of the Lattice Boltzmann Method for non-Newtonian and Free-surface Flow towards Applications in Wellbore Cementing

### 2018-01-03 response-to-reviewers.tex

\section*{Response to reviewers}

Thank you very much for your assistance in reviewing this paper. We are sure that with your help, this has become a much better work. We hope that the attached edited copy of ``Numerical Investigation of the Accuracy, Stability, and Efficiency of Lattice Boltzmann Methods in Simulating Non-Newtonian Flow'' will answer any questions previously raised, and will be satisfactory for publication.

The following is a list of all comments received, and how they have been addressed in the updated version:

  \subsection*{Reviewer 1}
	\begin{enumerate}
		\item \emph{The authors may test other non-Newtonian fluids more than the present Bingham plastics and power-law fluids.}

		Thank you for your comment.
		More complex non-Newtonian behavior would indeed to be interesting to investigate.
		However, as some non-Newtonian constitutive models combine and/or modify Bingham plastic and power-law behavior (e.g. Herschel–Bulkley fluids) these two non-Newtonian fluids were seen as representing the building blocks, in a sense, of more complex non-Newtonian behavior.
		While investigating more non-Newtonian fluids presents interesting future work, we believe that, in the interest of brevity, the current work should remain focused on Bingham plastics and power-law fluids.

		\item \emph{How about the non-Newtonian flow in complex geometries,such as the porous media? It is more strict for the stability requirement.}

		Thank you for the suggestion.
		Similarly, although the authors agree that such an investigation would be interesting, in the interest of brevity we have chosen not to expand the current investigation beyond Poiseuille flow and lid-driven cavity flow.

		\item \emph{An outlook for further models to improve the simulation stability of LBM.}

		The discussion of future work at the end of the paper has been expanded to discuss more possible LBM models and stabilization techniques.

@@

	\item \emph{Page 7: ``Due to the discrete nature of velocity in LBM, the integrals simply become
		summations'' Please be more precise, it is an intrinsic part of the discretization
		procedure to replace continuous integrals (moments) by numerical quadrature, i.e.
		Gauss-Hermite quadrature in LBM.}

	The aforementioned statement was removed and replaced by explicitly noting that the summations are a Gaussian-type quadrature.
	This was followed by relating the weights of the quadrature to the equilibrium distribution function.
	Additionally, the following connections to numerical quadrature are made directly following Eq. (7) by: (1) mentioning how the weights in Eq. (7) are determined, (2) the relationship between the weights in Eq. (7) and the quadrature weights are given and (3) the reader is referred to some additional reading if they are interested in further detail regarding the numerical quadrature.
	Thank you for this comment.
	The changes mentioned above should help improve the clarity of this paper.

	\item \emph{Page 4/15: You mention that computing the strain rate tensor is second order
		accurate, but the reviewer couldn’t find any comment on the accuracy of LBM to
		Navier-Stokes. This could be mentioned in the context with Chapman-Enskog
		expansion (p. 8)}

	In the second to last paragraph of page 6, it is mentioned that in the limit of small Mach number and small Knudsen number that the LBE recovers the Navier-Stokes equations.
	Also, the last paragraph of Section 3.1 discusses some of the nuances of the simulation Mach number in the context of simulating non-Newtonian fluids.
	In particular, it mentions that, although reducing the Mach number should increase the accuracy of the LBE with respect to the Navier-Stokes equations, it can introduce error in the constitutive relation.

@@

	\item \emph{Page 15: An exemplary figure (e.g. for $\tau_y = 4.0*10^-5$) comparing the viscosity
		according to the exact Bingham relationship and the approximations according to
		Papanastasiou with $m=10^5$ and $m=10^8$, respectively, would allow a comparison at
		first glance. Also the reader could see differences directly.}

	This figure was added.
	Thank you for the suggestion.

	\item \emph{Page 22: Why don’t you compare the MRT calculations to the BGK,
		$m=10^8$, cases, since $m=10^8$ was used for MRT as well?}

	This is a good point.
	The discussed has been expanded to compare the MRT calculations to the BGK, $m = 10^8$ cases.

@@

	\item \emph{Page 23, 24: You draw some conclusions between the ``ghost'' modes ($\epsilon$ and $q$) and
		nonphysical oscillations. For example you state (page 23): ``that the nonphysical
		oscillations the BGK with $m = 10^8$ displayed was in part due to the $\epsilon$ moment''.
		To the reviewer there is no direct proof visible for these conclusions. It might also be
		the other way around; oscillations in the distributions cause the oscillations in the
		moments. Please explain this in more detail.}

	The authors agree with the reviewer.
	The evidence is inconclusive and as such one can only make loose inferences on the actual cause of oscillations.
	The language in the discussion in pages 23-29 has been altered to reflect the fact that the results do not offer direct proof.
	In addition, parts of the discussion has been expanded to better motivate and explain why the authors believe the nonequilibrium $\epsilon$ and $q_x$ moments may be the primary source of oscillations.

@@

	\item \emph{Page 23, 24: Did you investigate if there are oscillations in the moment e, related to
		the bulk viscosity? It is well known that a high bulk viscosity is stabilizing and thus
		producing some of the superior MRT stability properties.}

	The nonequilibrium $e$ moment was indeed calculated and investigated.
	It has been added to the discussion of the nonequilibrium moments.

@@

	\item \emph{Page 25: ``...ghost mode associated with the q x moment dominates and is the
		primary source of oscillations.'' Again, this is not obvious for the reviewer, please
		prove this conclusion.}

	Again, the authors agree that direct proof cannot be found in the results.
	The language has been changed to reflect this.
	In addition, parts of the discussion has been expanded to better motivate and explain why the authors believe the nonequilibrium $\epsilon$ and $q_x$ moments may be the primary source of oscillations.

@@

	\item \emph{Page 25: ``Overall, the results presented suggest that if an LBM collision scheme is to
		be developed ... , it should focus on a means of dampening the nonequilibrium
		energy flux moments, namely q x and q y''. Did you consider the two relaxation times
		model (TRT) of Ginzburg et al.?}

	The two relaxation time model was not investigating. However, the discussion of possible future work has been expanded to included the TRT.

@@

	\item \emph{Page 27: Lid driven cavity: Please explain the setup of the simulation in more detail:
		e.g. boundary conditions, lattice Mach number, ...}

	The setup of the simulations is given in more detail.

	\item \emph{Page 30: Some exemplary figures of the flow pattern in the cavity, e.g. streamlines or
		contour plots, would give more insight, especially about smoothness or oscillations in
		the solution}

	Streamlines comparing the flow patterns of the various collision schemes for some of the more interesting non-Newtonian fluids and Reynolds numbers are now provided and accompanied by discussions of the results.
	Thank you for the suggestion.
	This should give a clearer picture of the similarities and differences between the solutions of the various collision schemes.

### 2018-01-04 src/stability-paper.tex
finished responses to reviewers
addressed Scott's comments

The Lattice Boltzmann method (LBM) is a numerical method based on computational statistical mechanics that is well-suited for approximating complex flow behaviors such as non-Newtonian, free surface, and multiphase multicomponent flow.
LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it, along with the equilibrium distribution function, determines the physics that are simulated (e.g. constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of their accuracy, numerical stability and computational efficiency.
The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater than the BGK collision operator.
It was also shown that, although the cost of artificial dissipation is that it introduces error in the constitutive response of the fluid (and therefore, may not accurately capture the physics of the flow), it can be an effective technique for stabilizing the numerics of non-Newtonian, lid-driven cavity flow simulations, and is particularly effective for stabilizing shear-thinning fluids.

@@

The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}.
The lattice is two-dimensional with nine discrete velocities at each node.
There is a stationary particle, there are four discrete velocities of magnitude 1, $\begin{Bmatrix}1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & -1\end{Bmatrix}^T$, and there are four discrete velocities of magnitude $\sqrt{2}$, $\begin{Bmatrix}1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & -1\end{Bmatrix}^T$, $\begin{Bmatrix}1 & -1\end{Bmatrix}^T$.
The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as:
\begin{equation}
f_i(\pos + \pvel_i \Delta t, t + \Delta t) = f_i(\pos, t) + \colop_i(\pos, t),
\end{equation}
\noindent where, for D2Q9, $i = 0, 1, ..., 8$ is the index of the discrete velocity vector.
\begin{figure}
  \centering
  \includegraphics[width=10cm]{figs/d2q9}
  \caption{Schematic of D2Q9 lattice, each node is connected to its neighbor by one of eight discrete velocity vectors}
  \label{fig:d2q9}
\end{figure}
Using Chapman-Enskog multiscale analysis, one can show that in the limit of small Mach number and small Knudsen number (i.e. the ratio of the mean free path to the characteristic length of the flow) the LBE recovers the Navier-Stokes equations.

@@

The challenge associated with high viscosity fluids is that particular distributions may never relax as ``far'' toward quasiequilibrium as it is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in extreme underrelaxation to the point of being negligible.
Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the local collision frequency may result in overrelaxation in certain parts of the domain and extreme underrelaxation in others.

@@

To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
The idea of artificial dissipation is to increase numerical stability while sacrificing some physical accuracy.
%Lyapanov stability is the idea that solutions at equilibrium stay there.
A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution.
From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation is necessary?'' and ``how much artificial dissipation does one introduce when it is necessary?''.

@@

The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing}.
The Reynold's number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution.
The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
Because the accuracy of LBM in recovering the Navier-Stokes equations depends on the Mach number being small, the Mach number of each simulation is also tabulated.

\begin{table}
\centering
\caption{Bingham plastic Poiseuille flow}
\vspace{0.5cm}
\begin{tabulary}{\linewidth}{l l r r r r r r r r}
\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Ma$ & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
\hline \\
BGK & No & $10^5$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0204 & 0.0411 & 2329 \\
& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0503 & 0.0891 & 3345 \\
& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.1161 & 0.1879 & 2029 \\
\\
BGK & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0330 & 0.0670 & 3509 \\
& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0788 & 0.1570 & 4838 \\
& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.1991 & 0.3539 & 3790 \\
\\
BGK & Yes & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0361 & 0.0823 & 3567 \\
& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.2439 & 0.3832 & 4800 \\
& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.7533 & 1.0718 & 3507 \\
\\
MRT-1 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0018 & 0.0018 & 7908 \\
& & & 12.0 & 3.04 & 4.04 & 0.016 &  0.0026 & 0.0026 & 7160 \\
& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.0041 & 0.0041 & 5559 \\
\\
MRT-2 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0012 & 0.0012 & 5284 \\
& & & 8.0 & 4.42 & 1.85 & 0.024 & 0.0017 & 0.0017 & 5290 \\
& & & 12.0 & 3.04 & 4.04 & 0.016 & 0.0024 & 0.0024 & 5244 \\
& & & 16.0 & 1.92 & 8.52 & 0.010 & 0.0038 & 0.0038 & 4532 \\
\\
\label{tab:poise-bing}
\end{tabulary}
\end{table}

@@

On average, the LBM model using the MRT-1 collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$; and the model using MRT-2 took 2.1 times more computing time than the BGK collision operator with $m = 10^5$.
The increased computing for the MRT-1 and MRT-2 with respect to the BGK collision operator with $m = 10^8$ was approximately 2.2 and 1.7, respectively, for the smaller yield stresses ($\tau_y = 4 \times 10^-5, 8 \times 10^-5$).
For larger yield stresses ($\tau_y = 12 \times 10^-5, 16 \times 10^-5$), the increased computing time was less, namely 1.5 and 1.1, respectively.
The increased computational efficiency of the BGK collision operator with $m = 10^8$ with respect to its MRT counterparts was less for the larger yield stresses, likely because the nonphysical oscillations in the BGK solutions resulted in requiring more iterations for the constitutive relation to converge.
However, the increased computing time can be justified for the MRT collision operator because the MRT solutions did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e. when $m = 10^8$. 
The MRT collision schemes were, therefore, the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT-1 model.
In contrast to the BGK with $m = 10^8$, there were no significant nonphysical oscillations in the quasiequilibrium distributions.

@@

Recall that: (1) when the relaxation rates $s_0 = s_1 = ... = s_8 = \omega$, the MRT collision operator is equivalent to the BGK collision operator (where $\omega$ is the collision frequency in the BGK sense), and (2) that the relaxation rates $s_1, s_2, s_4$ and $s_6$ are chosen with flexibility and may be tuned in an effort to increase stability.
Thus, an important question to ask is, if the strength of the MRT collision operator is damping out the nonequilibrium velocity moments with tunable relaxation parameters, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of these nonequilibrium moments and if so which ones?
\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\epsilon^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
\begin{equation}
\frac{||\boldsymbol{\epsilon} - \boldsymbol{\epsilon}^{eq}||_2}{||\boldsymbol{\epsilon}^{eq}||_2}
\end{equation}
\noindent where $||.||_2$ is the Euclidean norm, $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e. $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT-1 or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed may in part be due to the $\epsilon$ moment.

@@

It is also possible that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time (i.e. oscillations build in amplitude with time).
\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated, respectively, as:
\begin{equation}
  \int_0^{T} \frac{||\boldsymbol{\epsilon}(t) - \boldsymbol{\epsilon}^{eq}(t)||_2}{||\boldsymbol{\epsilon}^{eq}(t)||_2} dt,
\end{equation}
\noindent and,
\begin{equation}
  \int_0^{T} \frac{||\mathbf{q_x}(t) - \mathbf{q_x}^{eq}(t)||_2}{||\mathbf{q_x}^{eq}(t)||_2} dt,
\end{equation}
\noindent where $T$ is the current time. 
\Fref{fig:qx} and \Fref{fig:qx-cumulative} show that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the nonequilibrium $q_x$ moment may be the dominated source of oscillations.
It can be inferred that the nonequilibrium $q_x$ moment was much larger than the nonequilibrium $q_y$ moment in this case because the Poiseuille flow was in the x-direction.
Overall, the results presented suggest that if an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it may be important to focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.

@@

Lastly, simulations with the same collision schemes, Reynold's numbers, and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynold's numbers and Bingham numbers in \Fref{tab:poise-bing}.
The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing-2}.

@@

It is also interesting that, despite the finer grid resolution for the same characteristic flow, the MRT-1 and MRT-2 collision operators were less accurate on the $32 \times 128$ lattices than the $32 \times 64$ lattices.
A possible explanation for this nonintuitive result is that, as mentioned in~\cite{conrad2015accuracy}, although non-Newtonian lattice Boltzmann simulations are second order accurate, the accuracy is not preserved for every simulation Mach number.
Although the LBE recovers the Navier-Stokes equations in the limit of small Mach number, an additional artificial viscosity can be introduced for certain Mach numbers, and the optimal simulation Mach number depends on the constitutive parameters of the non-Newtonian fluid and the collision operator.
Thus, although the error in the LBE with respect to the Navier-Stokes equations was reduced by lowering the Mach number, an increase in the error in the constitutive relation may have been introduced that caused the MRT-1 and MRT-2 $32 \times 128$ simulations to have a greater net error (with respect to the analytical solution) than their $32 \times 64$ counterparts.

>>>>

The Lattice Boltzmann method (LBM) is a numerical method based on computational statistical mechanics that is well-suited for approximating complex flow behaviors such as non-Newtonian, free surface, and multiphase multicomponent flow.
LBM is typically applied to simulate flow through a series of time steps, each consisting of streaming particle distributions to neighboring nodes and collisions of particle distributions at each node through a collision operator.
The collision operator is of interest because it, along with the equilibrium distribution function, determines the physics that are simulated (e.g. constitutive laws, interfacial dynamics, etc.) and it has implications on numerical stability and computational efficiency.
This work examines various collision operators and methods for stability enhancement for their suitability for simulating non-Newtonian fluid flows in terms of their accuracy, numerical stability and computational efficiency.
The investigation was carried out as a numerical study looking for qualitative, yet practical, results, including testing the BGK and MRT collision operators, with and without entropic filtering, as applied to Bingham plastics and power-law fluids.
Two different benchmark problems were chosen for the flows: Poiseuille flow, and lid-driven square cavity flow.
The results of the numerical study showed that the MRT collision operator can have an advantage in terms of stability and accuracy for a variety of non-Newtonian flow behaviors, but at an increased computing cost that was, in some cases, as much as five times greater than the BGK collision operator.
It was also shown that, although it introduces error in the constitutive response of the fluid (and therefore, may not accurately capture the physics of the flow), artificial dissipation can be an effective technique for stabilizing the numerics of non-Newtonian, lid-driven cavity flow simulations, and is particularly effective for stabilizing shear-thinning fluids.

@@

The D2Q9 lattice was used in the current work (shown in \Fref{fig:d2q9}), which is commonly used for two-dimensional, incompressible flow simulations~\cite{Suc01}.
The lattice is two-dimensional with nine discrete velocities at each node.
There is a stationary particle, there are four discrete velocities of magnitude 1, $\begin{Bmatrix}1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 0\end{Bmatrix}^T$, $\begin{Bmatrix}0 & -1\end{Bmatrix}^T$, and there are four discrete velocities of magnitude $\sqrt{2}$, $\begin{Bmatrix}1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & 1\end{Bmatrix}^T$, $\begin{Bmatrix}-1 & -1\end{Bmatrix}^T$, $\begin{Bmatrix}1 & -1\end{Bmatrix}^T$.
The discretized version of the Boltzmann equation, or the lattice Boltzmann equation (LBE), is given as:
\begin{equation}
f_i(\pos + \pvel_i \Delta t, t + \Delta t) = f_i(\pos, t) + \colop_i(\pos, t),
\end{equation}
\noindent where, for D2Q9, $i = 0, 1, ..., 8$ is the index of the discrete velocity vector.
\begin{figure}
  \centering
  \includegraphics[width=10cm]{figs/d2q9}
  \caption{Schematic of D2Q9 lattice, each node is connected to its neighbor by one of eight discrete velocity vectors}
  \label{fig:d2q9}
\end{figure}
Using Chapman-Enskog multiscale analysis, one can show that, in the limit of small Mach number and small Knudsen number (i.e. the ratio of the mean free path to the characteristic length of the flow), the LBE recovers the Navier-Stokes equations.

@@

The challenge associated with high viscosity fluids is that some distributions may never relax as ``far'' toward quasiequilibrium as is physical because $\omega \rightarrow 0$ as $\nu \rightarrow \infty$, resulting in underrelaxation.
Concerns with sharp gradients, overrelaxation, and underrelaxation are particularly relevant in non-Newtonian flow because of the nonlinear constitutive relationship between shear stress and strain-rate.
The nonlinear constitutive relationship can lead to sharp gradients in $\rho$ or $\mvel$, and depending on the form of the function $\mu_{app}(\dot{\gamma})$, the local collision frequency may result in overrelaxation in certain parts of the domain and underrelaxation in others.

@@

To reduce nonequilibrium fluctuations in LBM, one can introduce artificial dissipation.
The idea of artificial dissipation is to increase numerical stability while sacrificing some physical accuracy.
%Lyapanov stability is the idea that solutions at equilibrium stay there.
A practical goal then would be to use only the necessary amount of artificial dissipation in order to ensure a stable solution within a reasonable amount of error in the constitutive relation.
From this goal two questions naturally arise: ``under what criteria does one decide that artificial dissipation should be introduced?'' and ``how much artificial dissipation should be introduced?''.

@@

The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing}.
The Reynolds number was computed by $Re = \frac{\rho U H}{\mu_p}$, where $U$ is the maximum velocity given by the analytical solution.
The Bingham number was computed by $Bn = \frac{\tau_y H}{\mu_p U}$.
Because the accuracy of LBM in recovering the Navier-Stokes equations depends on the Mach number being small, the Mach number of each simulation is also tabulated.

\begin{table}
\centering
\caption{Bingham plastic Poiseuille flow}
\vspace{0.5cm}
\begin{tabulary}{\linewidth}{l l r r r r r r r r}
\pbox{20cm}{Collision \\ Operator} & \pbox{20cm}{Median \\ Filter} & $m$ & \pbox{20cm}{$\tau_y$ \\ $(\times 10^{-5})$} & $Ma$ & $Re$ & $Bn$ & $L_2$ & $L_\infty$ & \pbox{20cm}{Time \\ (sec)} \\
\hline \\
BGK & No & $10^5$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0062 & 0.0153 & 1857 \\
& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0204 & 0.0411 & 2329 \\
& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0503 & 0.0891 & 3345 \\
& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.1161 & 0.1879 & 2029 \\
\\
BGK & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0109 & 0.0282 & 2831 \\
& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0330 & 0.0670 & 3509 \\
& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0788 & 0.1570 & 4838 \\
& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.1991 & 0.3539 & 3790 \\
\\
BGK & Yes & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0100 & 0.0273 & 2903 \\
& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0361 & 0.0823 & 3567 \\
& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.2439 & 0.3832 & 4800 \\
& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.7533 & 1.0718 & 3507 \\
\\
MRT-1 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0013 & 0.0013 & 5914 \\
& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0018 & 0.0018 & 7908 \\
& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0026 & 0.0026 & 7160 \\
& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.0041 & 0.0041 & 5559 \\
\\
MRT-2 & No & $10^8$ & 4.0 & 0.033 & 6.05 & 0.68 & 0.0012 & 0.0012 & 5284 \\
& & & 8.0 & 0.024 & 4.42 & 1.85 & 0.0017 & 0.0017 & 5290 \\
& & & 12.0 & 0.016 & 3.04 & 4.04 & 0.0024 & 0.0024 & 5244 \\
& & & 16.0 & 0.010 & 1.92 & 8.52 & 0.0038 & 0.0038 & 4532 \\
\\
\label{tab:poise-bing}
\end{tabulary}
\end{table}

@@

On average, the LBM model using the MRT-1 collision operator took 2.9 times more computing time than the BGK collision operator with $m = 10^5$; and the model using MRT-2 took 2.1 times more computing time than the BGK collision operator with $m = 10^5$.
The increased computing time for the MRT-1 and MRT-2 with respect to the BGK collision operator with $m = 10^8$ was approximately 2.2 and 1.7, respectively, for the smaller yield stresses ($\tau_y = 4 \times 10^{-5}, 8 \times 10^{-5}$).
For the larger yield stresses ($\tau_y = 12 \times 10^{-5}, 16 \times 10^{-5}$), the increased computing time was less, namely 1.5 and 1.1, respectively.
The increased computational efficiency of the BGK collision operator with $m = 10^8$ with respect to its MRT counterparts was less for the larger yield stresses, likely because the nonphysical oscillations in the BGK solutions required more iterations for the constitutive relation to converge.
However, the increased computing time can be justified for the MRT collision operator because the MRT solutions did not suffer from the same level of nonphysical oscillations as the BGK collision operator when using the more accurate approximation of the Bingham plastic model, i.e. when $m = 10^8$. 
The MRT collision schemes were, therefore, the most accurate solutions for all cases; and because the difference in accuracy between the MRT-1 and MRT-2 collision schemes was negligible, the remainder of the discussion in this section will focus on comparing the BGK collision schemes and the MRT-1 collision scheme.
\Fref{fig:feq-vs-f_mrt} compares particle distribution, $f_i$, profiles to their respective quasiequilibrium profiles, $f_i^{eq}$ for the MRT-1 model.
In contrast to the BGK with $m = 10^8$, there were no significant nonphysical oscillations in the quasiequilibrium distributions.

@@

Recall that: (1) when the relaxation rates $s_0 = s_1 = ... = s_8 = \omega$, the MRT collision operator is equivalent to the BGK collision operator (where $\omega$ is the collision frequency in the BGK sense), and (2) that the relaxation rates $s_1, s_2, s_4$ and $s_6$ are chosen with flexibility and may be tuned in an effort to increase stability.
Thus, an important question to ask is, if the strength of the MRT collision operator is damping out the nonequilibrium velocity moments with tunable relaxation parameters, then are the nonphysical oscillations for the BGK with $m = 10^8$ (with and without median filtering) a result of these nonequilibrium moments and, if so, which ones?
\Fref{fig:epsilon} shows a measure of the nonequilibrium $\epsilon$ moment, $\epsilon^{neq}$, with respect to time for each of the collision schemes, which was calculated as:
\begin{equation}
\frac{||\boldsymbol{\epsilon} - \boldsymbol{\epsilon}^{eq}||_2}{||\boldsymbol{\epsilon}^{eq}||_2}
\end{equation}
\noindent where $||.||_2$ is the Euclidean norm, $\epsilon_j = 4f_0(\pos_j) - 2f_1(\pos_j) - 2f_2(\pos_j) - 2f_3(\pos_j) - 2f_4(\pos_j) + f_5(\pos_j) + f_6(\pos_j) + f_7(\pos_j) + f_8(\pos_j)$ is related to the square of the energy, and $\pos_j = \begin{Bmatrix}16 & y_j\end{Bmatrix}^T$ for $j = 1, 2, ..., 64$, i.e. $\pos_j$ is taken at 16 nodes in from the left in the x-direction and for the full height of the channel in the y-direction.
The values measured for the BGK with $m = 10^8$, with and without median filtering, were significantly higher than for either the MRT-1 or BGK with $m = 10^5$ collision schemes, which suggests that the nonphysical oscillations the BGK with $m = 10^8$ displayed may in part be due to the $\epsilon$ moment.

@@

It is also possible that the nonequilibrium $\epsilon$ and $q_x$ moments cause nonphysical oscillations in particle distributions, $f_i$, with a cumulative effect in time (i.e. oscillations build in amplitude with time).
\Fref{fig:epsilon-cumulative} and \Fref{fig:qx-cumulative} show the cumulative measures of the nonequilibrium moments in time, which was calculated, respectively, as:
\begin{equation}
  \int_0^{T} \frac{||\boldsymbol{\epsilon}(t) - \boldsymbol{\epsilon}^{eq}(t)||_2}{||\boldsymbol{\epsilon}^{eq}(t)||_2} dt,
\end{equation}
\noindent and,
\begin{equation}
  \int_0^{T} \frac{||\mathbf{q_x}(t) - \mathbf{q_x}^{eq}(t)||_2}{||\mathbf{q_x}^{eq}(t)||_2} dt,
\end{equation}
\noindent where $T$ is the current time. 
\Fref{fig:qx} and \Fref{fig:qx-cumulative} show that both the peak and cumulative values of the relative $L_2$ norm of $\mathbf{q_x}^{neq}$ are much larger than the peak and cumulative values of the relative $L_2$ norm of $\boldsymbol{\epsilon}^{neq}$, which suggests that the nonequilibrium $q_x$ moment may be the dominant source of oscillations.
It can be inferred that the nonequilibrium $q_x$ moment was much larger than the nonequilibrium $q_y$ moment in this case because the Poiseuille flow was in the x-direction.
Overall, the results presented suggest that if an LBM collision scheme is to be developed for simulating high yield stress, Poiseuille-type flow that is more accurate than the BGK with $m = 10^5$ and more computationally efficient than the MRT, it may be important to focus on a means of dampening the nonequilibrium energy flux moments, namely $q_x$ and $q_y$.

@@

Lastly, simulations with the same collision schemes, Reynolds numbers, and Bingham numbers were run on $128 \times 32$ lattices in order to investigate the effect of grid resolution.
The plastic viscosity was held constant ($\mu_p = 0.2$.), while the yield stress and pressure gradient were varied in order to match the Reynolds numbers and Bingham numbers in \Fref{tab:poise-bing}.
The relative $L_2$ error, relative $L_{\infty}$ error, and computation time for each simulation are presented in \Fref{tab:poise-bing-2}.

@@

It is also worth noting that, despite the finer grid resolution for the same characteristic flow, the MRT-1 and MRT-2 collision operators were less accurate on the $32 \times 128$ lattices than the $32 \times 64$ lattices.
A possible explanation for this nonintuitive result is that, as mentioned in~\cite{conrad2015accuracy}, although non-Newtonian lattice Boltzmann simulations are second order accurate, the accuracy is not preserved for every simulation Mach number.
Although the LBE recovers the Navier-Stokes equations in the limit of small Mach number, an additional artificial viscosity can be introduced for certain Mach numbers, and the optimal simulation Mach number depends on the constitutive parameters of the non-Newtonian fluid and the collision operator.
Thus, although the error in the LBE with respect to the Navier-Stokes equations was reduced by lowering the Mach number, an increase in the error in the constitutive relation may have been introduced that caused the MRT-1 and MRT-2 $32 \times 128$ simulations to have a greater net error (with respect to the analytical solution) than their $32 \times 64$ counterparts.
